---
title: "Untitled"
author: "STA303 In Class"
date: '2019-10-25'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this tutorial we will learn how to use `R` for basic data anlaysis. We are
going to 

1. Do some basic scientific computing in `R`.
1. Do some basic data analysis in `R`.
1. Export the results as a document using `RMarkdown`.

We will see examples of concepts and then you will complete short exercises on
your own.

You must download the `tidyverse` before starting. Type `install.packages("tidyverse")`
and wait a while. Then type

```{r tidyverse-1}
library(tidyverse)
```

to load the package so you can use its functions.

# Scientific Computing

Scientific computing refers to the use of a computer to do complicated calculations.
This is useful to you for your studies and also is a good way to get introduced 
to `R`.

As a motivating example, we are going to compute the $2^{nd}$-order Taylor expansion
of a simple trigonometric function. This will introduce you to the following
concepts:

1. Variables and assignment,
1. functions, and
1. plots

## Problem Setup

Recall from calclus that a sufficiently smooth function $f(x)$ can be written
as a power series:
\begin{equation}
f(x) = \sum_{k=0}^{\infty}\frac{f^{(k)}(x_{0})(x - x_{0})^{k}}{k!}
\end{equation}
where $x_{0}$ is a reference point and $f^{(k)}$ is the $k^{th}$ derivative
of $f$. This representation is exact when the sum is infinite.

However, you can truncate the sum at some finite $N$ to get a polynomial approximation
to $f$, called a *Taylor series approximation*. It is common to take $N = 1$ or $2$, to get a *linear* or *quadratic*
approximation. These are used all the time in applied science, and extensively
in statistics.

For our purposes, let's take $f(x) = \cos(x)$ and $x_{0} = 0$. Its derivatives are
\begin{equation}\begin{aligned}
f^{(1)}(x) &= -\sin(x) \\
f^{(2)}(x) &= -\cos(x) \\
\end{aligned}\end{equation}

How can we use `R` to compute the Taylor series approximation to $\cos(x)$? We need to

1. Tell `R` about the initial point $x_{0}$,
1. tell `R` about $f(x)$ and its first two derivatives,
1. tell `R` how to compute the value of the approximation for any $x$, and
1. produce a graphic showing the function and its approximation.

## Variables and assignment

### Basic concepts

To tell `R` about the initial point $x_{0} = 0$, we create a **variable** called
$x_{0}$ and **assign** the value $0$ to it. We can do this as follows:

```{r var-1}
# Create a variable simply by assigning it a value
x0 <- 0
```

This code uses the **assignment operatior** `<-` to create a **variable** `x0` and **assign** it value `0`. 

You can check that your variable was defined either by listing all the variables
currently defined:

```{r var-2}
ls()
```

or by specifically checking if the variable `x0` exists:

```{r var-3}
exists("x0")
```

or by typing the name of your variable and seeing what prints out:

```{r var-6}
x0
```

This last way also tells you the *value* of the variable.

**Exercise**: create a variable called `y` which takes the value $2$. Check that
it exists.

### Vectors

So what if you had more than one number you wanted to store? You could create
more than one variable, like `y1 <- 1; y2 <- 2`, but this becomes cumbersome
quickly. Instead, you can store many numbers in a *single variable*, by using
a **vector**. The `c()` (for "concatenate") function in `R` is used for this:

```{r var-4}
y1 <- c(1,2,3,4,5) # Create a vector 
y1 # See what it looks like
y1[1]
y1[2] # Access individual elements
y1[3:5] # Access a subset of elements
length(y1) # How many elements?
```

You can also assign values to existing variables:

```{r var-5}
y1 # Old value
y1 <- c(6,7,8) # Completely new value
y1
y1[2] <- 10 # Can assign to a subset
y1
```

**Exercise**: create a vector containing the values `1:10`. Change the fifth 
element to `20`. Print it out and make sure it's correct.

## Functions

Now that you can store numbers in `R`, you want to be able to do things to those
numbers. You do this using **functions**.

Functions are actually themselves "values" that you assign to variables, in the
same way as numbers and vectors. And to create a function, you use the `function()`
function. Confused? Check it out:

```{r fun-1}
# Create a function to multiply a number by 2:
multiply_by_2 <- function(x) {
  x * 2
}
# Call your function:
multiply_by_2(2)
# Call your function on a pre-defined variable:
multiply_by_2(x0)
# Many operations also work on vectors:
multiply_by_2(y1)
```

That's it! There are a few things to note:

1. Functions take zero or more **arguments**. Here there is one argument called
`x`. You can add more arguments or have no arguments. Arguments can be any type
of `R` object, and unlike other programming languages you don't have to pre-specify
what they are.

1. Functions may or may not return a value. If the last statement in a function
is an expression like `x * 2`, then the function returns that value. You can 
also use `return()` to return a value.

1. Functions can call other functions. In this example, `*` is itself a function!

Let's create functions to represent $f(x)$ and its first two derivatives:

```{r fun-2}
f <- function(x) cos(x)
f1 <- function(x) -sin(x)
f2 <- function(x) -cos(x)
f(c(-pi/2,0,pi/2)) # pi is a variable provided by R
f1(c(-pi/2,0,pi/2))
f2(c(-pi/2,0,pi/2))
```

**Exercise**: create a function that adds `2` to a number.

**Exercise**: create a function $g(x)$ that computes the tangent of an angle,
$\tan(x) = \sin(x)/\cos(x)$. What happens if you call $g(\pi/2)$?

## Putting it together: the Taylor expansion of $\sin(x)$

Now let's use our new knowledge of functions and vectors to compute the
Taylor expansion of $f(x) = \cos(x)$ to second order.

```{r taylor-1}
# Create a function that computes the Taylor expansion of f(x) at point x0
taylorfunction <- function(x) {
  f(x0) + f1(x0) * (x - x0) + (1/2)*f2(x0) * (x - x0)^2
}
# Try it out
f(0)
taylorfunction(0)
f(0.5)
taylorfunction(0.5)
# and so on
```

**Exercise**: $f^{(3)}(x) = -cos(x)$. Can you add a third order of approximation
to the above? Create a new function `taylorfunction3rd` that approximates $f(x)$
to third-order, and compare its values at $x = -.5,0,.5$ to $f(x)$ and `taylorfunction`.

## Creating graphics

We should assess our approximation visually. To do this we will use `ggplot2` to
plot $f(x)$ and its approximation.

`ggplot2` is a layer-based graphics system. It's best to just learn it by trial-and-error,
by seeing dozens or hundreds of examples and eventually figuring out how to search
for what you want to do. That's how I learned it anyways. But I'll try and explain.

You build up a plot in layers. First you add data; then you map variables to aesthetic
elements of the plot like the "x-axis" or the "colour"; then you tell it what to
draw (points, lines, bars, etc); then you annotate the plot with axis labels and 
other things. It looks like this:

```{r gg-1}
# First, start by saving a blank plot to a variable:
plt <- tibble(x = c(-5,5)) %>% # Give (-5,5) as the data; this tells ggplot where to draw the function to and from
  ggplot(aes(x = x)) # aes(): tell ggplot that the variable "x" goes on the x-axis.
# What does this look like?
plt
# Not much. Add a blank theme:
plt <- plt + theme_classic()
plt
# Now add a line for the true function f(x)
plt <- plt + geom_line(stat = "function",fun = f) # stat = "function" tells geom_line() to draw a function
plt
# Now add a line for the Taylor approximation
plt <- plt + geom_line(stat = "function",fun = taylorfunction,colour = "blue",linetype = "dotdash")
plt
# Finally, add some axis labels and a title
plt <- plt + labs(title = "Taylor approximation (blue) to f(x) = cos(x) (black) about x0 = 0",
                  x = "x",y = "f(x) and approximation")
plt
```

You don't have to do everything step-by-step like that in practice. You can do
it all at once:

```{r gg-2}
tibble(x = c(-5,5)) %>%
  ggplot(aes(x = x)) +
  theme_classic() +
  geom_line(stat = "function",fun = f) +
  geom_line(stat = "function",fun = taylorfunction,colour = "blue",linetype = "dotdash") +
  labs(title = "Taylor approximation (blue) to f(x) = cos(x) (black) about x0 = 0",
                  x = "x",y = "f(x) and approximation") +
  coord_cartesian(ylim = c(-1,1))
```

Notice how I set the y-limits of the plot to better capture the range I wanted to see.

**Exercise**: add your `taylor3rdorder` function to the above plot. Change the colour
to `"red"` or your favourite alternative.

# Data Anlaysis

`R` is a programming language that is optimized for interactive data analysis.
In this section of the tutorial we are going to analyze data using `R`.

We are going to analyze the "Old Faithful" Geyser data, a famous dataset on
eruption times from a famous geyser in Yellowstone National Park in Wyoming.
The data is actually available in `R` but for illustration purposes I saved
it to a text file which you can get directly from my github.

## Read in the data

The data is in "comma-separated value" format,
which means it's a text file where variables are separated by commas. Look at
the raw data [here](https://raw.githubusercontent.com/awstringer1/ssu-r-workshop/master/data/faithful.txt).

Here is how you read in the data.

```{r faithful-1}
faithful <- readr::read_csv(
  file = "https://raw.githubusercontent.com/awstringer1/ssu-r-workshop/master/data/faithful.txt",
  col_names = TRUE,
  col_types = c("nn")
)
# Take a look at it:
glimpse(faithful)
```

The data has two variables, `eruptions` and `waiting`, and 272 rows.

## Summary Statistics

We can use `R` to analyze the data. Let's look at some basic summary statistics:
mean, median, min/max and quantiles. You get the mean of a vector of numbers in
`R` using the `mean()` function. You access one column of a dataframe by using
the `$` operator. We do this as follows:

```{r faithful-2}
mean(faithful$eruptions)
mean(faithful$waiting)
```

It looks like the mean waiting time is `r mean(faithful$waiting)` and the mean number of 
eruptions is `r mean(faithful$eruptions)`.

You can get a five-number summary using the `summary()` function:

```{r faithful-3}
summary(faithful$eruptions)
```

**Exercise**: compute the five-number summary for the `waiting` variable.

## Plots

We can also plot the data, just like we plotted our functions. We'll use `ggplot`
to make a scatterplot. The premise is the same: build the plot up in layers.

```{r faithful-4}
faithful %>% # Provide the data to ggplot() directly
  ggplot(aes(x = waiting,y = eruptions)) + # Tell ggplot to put waiting on the x-axis and eruptions on the y-axis
  theme_classic() +
  geom_point(pch = 21,colour = "black",fill = "orange")
```

**Exercise**: add a title and x/y axis labels to this plot. Remember the `labs()` function
from the Taylor series example?

We can also look at the individual variables' distributions. A common type of plot
for this is a **boxplot** which essentially plots the five-number summary of a variable.
We can make this in `ggplot` as follows:

```{r faithful-5}
faithful %>%
  ggplot(aes(y = waiting)) +
  theme_classic() +
  geom_boxplot(width = 1) +
  coord_cartesian(xlim = c(-3,3)) +
  theme(axis.text.x = element_blank(),axis.ticks.x = element_blank()) +
  labs(title = "Distribution of waiting times until eruption, Old Faithful Geyser",
       x = "",y = "Waiting Time")
```
